Supp mat: time series paper

### 2018-02-19 writing/manuscript_draft.tex
first update of short format version of the article

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \usepackage{lineno}
\usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}
\linenumbers

@@

\textbf{Minimum time required to detect population trends: the need for long-term monitoring programs}
\vspace{3 mm}

Easton R. White\(^{*}\)

\begin{footnotesize} Center for Population Biology, University of California, Davis, California 95616 USA

*eawhite@ucdavis.edu
\end{footnotesize}

\vspace{5 mm}

@@

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? We determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, we use simulation methods and examine 878 populations of vertebrate species. Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power. For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation. These results point to the importance of sampling populations over long periods of time. We argue that statistical power needs to be considered in monitoring program design and evaluation. Time series less than 15-20 years are likely under-powered and potentially misleading.
\end{abstract}

@@

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols and
Williams 2006). Scientists need to know the time series length required
to address a specific question. A short time series may lead to wrong
conclusions given large natural year-to-year variability (McCain,
Szewczyk, and Knight 2016). Managers need to know when action is needed
for a population. Therefore, managers must understand when population
trend over time is actually meaningful. For example, the IUCN Red List
Categories and Criteria suggest, under Criterion A2, a species qualifies
as vulnerable if it has experienced a 30\% decline over 10 years, or 3
generations (IUCN 2012). For both scientific and management questions,
because sampling is typically expensive, we also do not want to sample
for longer than is necessary. For example, Gerber, DeMaster, and Kareiva
(1999) investigated the minimum time series required to estimate
population growth of the endangered, but recovering, eastern North
Pacific gray whale (\emph{Eschrichtius robustus}). They used a long-term
census to retroactively determine the minimum time series required to
assess threat status. They found that only 11 years were needed, eight
years before the delisting decision was made. This highlights the
importance of estimating the minimum time series required as an earlier
decision would have saved time and money (Gerber, DeMaster, and Kareiva
1999).

@@

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Wagner,
Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
and Jonzen (2011) examined the optimal allocation of effort between
spatial and temporal replicates. Using simple populations models, they
found that the allocation of effort depends on environmental variation,
spatial and temporal autocorrelation, and observer error. Rueda-Cediel
et al. (2015) also used a modeling approach, but parameterized a model
specific for a threatened snail, \emph{Tasmaphena lamproides}. They
found that for this short-lived organism, 15 years was adequate to
assess long-term trends in abundance. Hatch (2003) examined seabird
populations and determined the minimum time required to achieve high
statistical power ranged between 11 and 69 years. However, these
studies, and other past work, have typically focused either on
theoretical aspects of monitoring design or focused on only a few
species.

We use both simulations and empirical time series to determine the
minimum number of years required to address several questions. We
estimate the minimum time series length required (\(T_{min}\)) to assess
long-term changes in abundance via simple linear regression. First, we
estimate \(T_{min}\) using a simulation approach. Then we examine 878
population time series to estimate \(T_{min}\). In the supplementary
material, we determine \(T_{min}\) for related ideas: calculating
long-term growth rates, using more complicated population models,
varying statistical level and power, and the use of generalized additive
models.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
approach as it has been described elsewhere. Essentially, we use a
population model and repetitively simulate it for a number of years.
This approach requires us to determine values for our model parameters
(e.g.~population variability). As an example, we can take the following
population model for population size \(N\) at time \(t\):

@@

where \(\epsilon\) is a normally-distributed random noise term with mean
\(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\) is
also the trend strength of the increase or decrease (i.e.~the rate of
increase). It is important to note that any population model could be
substituted for equation \ref{sim_model}, as we do in the supplementary
material (Figs. \ref{fig:min_time_dist_log_pop},
\ref{fig:shark_example}).

Statistical power is then the proportion of simulations that meet some
criteria. Here, our criteria is whether the slope parameter from linear
regression is significant at the 0.05 threshold with statistical power
of 0.8. Statistical power of 0.8 would indicate that, if there was a
true trend in abundance, we would have a 0.8 probability of detecting
the trend. We also tested the effect of varying both the significance
level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (\(t=40\)). It is
clear that statistical power increases quickly with increases in length
of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is
greater than 0.8 (the dotted line), that is the minimum time required
(\(T_{min}\)).

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015); they compared the predictability of growth rates among
populations. The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The data includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The data
also includes information on generation length and survey
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. All analyses were conducted in R (R Core Team 2016).

For a subset of populations (\(n =\) 547), we had information on
biological characteristics from another paper (Myhrvold et al. 2015),
including body size and generation length. All 547 populations were
birds. We examine how the minimum time required is related to these
biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 35 years would have 34 possible contiguous
subsamples of length 2, 34 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 35
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Next, we
run a linear regression for each subsampled time series. Then, we
determine the proportion of subsamples of a particular length that have
estimated slope coefficients which are statistically different from
zero. We only look at the proportion of samples where the long-term, or
``true'', time series also has a significant slope. This proportion is a
measure of statistical power. Lastly, we determine which subsample
length is required to achieve a certain threshold of statistical power
(0.8, Cohen (1992)). The minimum subsampled length that met these
criteria is the minimum time series length required (\(T_{min}\)).

In the supplementary material, we show how the same approach described
here can be used to determine the minimum time required to estimate a
population's geometric growth rate (Figs.
\ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also determine
the minimum time required to estimate long-term trends according to
generalized additive models, instead of the simple linear models used
here (Fig. \ref{fig:gam_example}).

@@

We determined the minimum time series length (\(T_{min}\)) required to
address a particular question of interest. What is the minimum time
series length required to determine, via linear regression, the
long-term population trend? Here, the minimum time series length
required had high enough statistical power (\(1-\beta\) greater than
0.8) for a set significance level (\(\alpha\)) of 0.05. It is also
possible to alter \(1-\beta\) and \(\alpha\). Predictably, as we
increased our level of power or decreased \(\alpha\), \(T_{min}\)
increased (Fig. \ref{fig:min_time_vs_alpha_beta}). We estimated
\(T_{min}\) using two approaches. We briefly describe results from the
simulation approach and then discuss our empirical approach.

@@

We constructed a general population model where the trend strength
(i.e.~slope coefficient) over time could be a model parameter. We then
ran simulated time series of different lengths. From these simulations
we determined the minimum time series length required to achieve a
certain level of statistical power. In line with past work (Gerrodette
1987), we found the \(T_{min}\) increases (i.e.~more time is required)
with decreases in trend strength and with increases in population
variability (Figs. \ref{fig:theoretical_approach}c,d).

We chose a simple model, but any other population model could be used
(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
specific model choice should be tailored to the population of interest.
We explored how the simulation approach could be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
More specifically, we determined the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). We found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
the simulation approach described above, the minimum time required for
the lemon shark population was strongly dependent on model parameters.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distributions of the minimum time required for populations from
four different biological classes. The minimum time required calculation
corresponds to a significance level of 0.05 and statistical power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (inter-annual
variability in population size), and (c) temporal lag-1
autocorrelation.\label{fig:correlates}}
\end{figure}

We examined a database of 878 separate population time series
representing 478 species. This database consists of vertebrate species
with a variety of life history characteristics (Fig.
\ref{fig:biological_correlates}). We limited our analyses to populations
with at least 35 years of continuous sampling. We then examined the
minimum time required to estimate long-term trends via linear
regression.

Across all the populations we examined, we found an average minimum time
series length required (\(T_{min}\)) of 16.5 (\(\sigma\)=8.4), with a
wide distribution (Fig. \ref{fig:min_time_dist}). Estimates of
\(T_{min}\) varied between biological class (Fig.
\ref{fig:min_time_dist}). Ray-finned fish (class Actinopterygii)
typically had estimates of \(T_{min}\) over 20 years. Birds (class Aves)
had a much wider distribution of \(T_{min}\), but usually required less
years of sampling. Differences between these classes were explained by
differences in inter-annual variability in population size (Fig.
\ref{fig:class}). We also examined a subset of populations where
life-history characteristics were known. None of these explanatory
variables were predictive of the minimum time series length required
(Fig. \ref{fig:biological_correlates}).

The minimum time series length required was strongly correlated with
trend strength, population variance, and autocorrelation (Fig.
\ref{fig:correlates}). This is in line based on our simulations and
those of others (Rhodes and Jonzen 2011). Using a generalized linear
model, with a Poisson error structure, all three of these explanatory
variables were significant and had large effect sizes (see Table A1).
Combined, trend strength, population variance, and autocorrelation
account for about 72.6\% of the explained deviance (Zuur et al. 2009) in
minimum time series length required.

Lastly, we tested model sensitivity by using generalized additive models
(GAMs) instead of simple linear regression. Again, we examined the
minimum time required to estimate long-term population trends (Fig.
\ref{fig:gam_example}). We found that although we obtain a similar
distribution of minimum times required for GAMs, the minimum time
required for GAMs is on average shorter than for linear regression (Fig.
\ref{fig:min_time_dist_gam}).

@@

We explored two approaches to estimate the minimum time series length
required to address a particular question of interest. We asked, what is
the minimum time series length required to determine long-term
population trends using linear regression? This is one of the simplest
questions one could ask of a time series. The simulation-based approach
has been suggested by others, especially in situations more complicated
than that suited for classic power analysis (Gerrodette 1987; P. C.
Johnson et al. 2015; Bolker 2008). Our simulations support past work
that longer time series are needed when the trend strength (i.e.~rate of
increase or decrease) is weak or when population variability is high
(Gerrodette 1987). We also showed how the simulation model can be
altered for a particular population (Fig. \ref{fig:shark_example}) or
question (Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).

Here, we focus on an empirical approach to estimate the minimum time
series length required to assess changes in abundance over time. We
examined 878 population time series (all longer than 35 years). We then
subsampled each to determine the minimum time required to achieve a
desired significance level and power for linear regression. Statistical
power is important as it provides on information as to the necessary
samples required to determine a significant trend (Legg and Nagy 2006).
We found that at least 15-20 years of continuous monitoring were
typically necessary (Fig. \ref{fig:min_time_dist}). This time frame is
in line with past work on a short-lived snail species (Rueda-Cediel et
al. 2015) and a long-lived whale species (Gerber, DeMaster, and Kareiva
1999). Hatch (2003) used seabird monitoring data to estimate minimum
sampling requirements. He found that the time required ranged from 11 to
69 years depending on species, trend strength, and study design.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{setspace}  
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \usepackage{lineno}
\usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}

@@

Title: Minimum time required to detect population trends: the need for
long-term monitoring programs \vspace{7 mm}

Author: \textsc{Easton R. White$^{1,2}$} \vspace{3 mm}

Address:\emph{
    \\$^1$Center for Population Biology \\
    University of California, Davis \\
    2320 Storer Hall \\
        University of California, Davis \\
        One Shields Avenue Davis, CA, USA} \vspace{3 mm}

\(^2\)Corresponding author:
\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}} , 1
480 203 7931 \vspace{3 mm}

Number of words (including 300 words per figure): 6,950 \vspace{1 mm}

Number of figures: 5 \vspace{1 mm}

Number of references: 34 \vspace{1 mm}

Supplementary material: 9 figures and 3 tables \vspace{3 mm}

Data and code for all the figures and tables can be found at
(\url{https://github.com/erwhite1/time-series-project}).

To be submitted to: \emph{Biological Conservation} \vspace{3 mm}

Available as a preprint at: \url{https://peerj.com/preprints/3168/}
\vspace{3 mm}

Color does not need to be used for any figures in print \vspace{3 mm}

\vspace{3 mm}

\begin{spacing}{2}

@@

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? I determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, I use simulation methods and examine 822 populations of vertebrate species. Here I show that on average 15.9 years of continuous monitoring are required in order to achieve a high level of statistical power. However, there is a wide distribution around this average, casting doubt on simple rules of thumb. For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation. However, there were no life-history traits (e.g. generation length) that were predictive of the minimum time required. These results point to the importance of sampling populations over long periods of time. I argue that statistical power needs to be considered in monitoring program design and evaluation. Short time series are likely under-powered and potentially misleading.
\end{abstract}

@@

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols and
Williams 2006). Scientists need to know the time series length required
to address a specific question. A short time series may lead to wrong
conclusions given large natural year-to-year variability (McCain,
Szewczyk, and Knight 2016). Managers need to know when action is needed
for a population. Therefore, managers must understand when population
trend over time is actually meaningful. For example, the International
Union for Conservation of Nature (IUCN) Red List Categories and Criteria
suggest, under Criterion A2, a species qualifies as vulnerable if it has
experienced a 30\% decline over 10 years, or three generations (IUCN
2012). For both scientific and management questions, because sampling is
typically expensive, we also do not want to sample for longer than is
necessary. For example, Gerber, DeMaster, and Kareiva (1999)
investigated the minimum time series required to estimate population
growth of the endangered, but recovering, eastern North Pacific gray
whale (\emph{Eschrichtius robustus}). They used a long-term census to
retroactively determine the minimum time series required to assess
threat status. They found that only 11 years were needed, eight years
before the delisting decision was made. This highlights the importance
of estimating the minimum time series required as an earlier decision
would have saved time and money (Gerber, DeMaster, and Kareiva 1999).
Further, waiting too long to make a decision can imperil a species where
management action could have been taken earlier (Martin et al. 2012;
Martin et al. 2017).

@@

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Wagner,
Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
and Jonzen (2011) examined the optimal allocation of effort between
spatial and temporal replicates. Using simple populations models, they
found that the allocation of effort depends on environmental variation,
spatial and temporal autocorrelation, and observer error. Rueda-Cediel
et al. (2015) also used a modeling approach, but parameterized a model
specific for a threatened snail, \emph{Tasmaphena lamproides}. They
found that for this short-lived organism, 15 years was adequate to
assess long-term trends in abundance. However, these studies, and other
past work, have typically focused either on theoretical aspects of
monitoring design or focused on only a few species.

I use both simulations and empirical time series to determine the
minimum number of years required to address several questions. I
estimate the minimum time series length required (\(T_{min}\)) to assess
long-term changes in abundance via simple linear regression. First, I
estimate \(T_{min}\) using a simulation approach. Then I examine 822
population time series to estimate \(T_{min}\). In the supplementary
material, I determine \(T_{min}\) for related ideas: using more
complicated population models, varying statistical level and power, and
the use of generalized additive models.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). I only briefly discuss this
approach as it has been described elsewhere. Essentially, a population
model is simulated repetitively for a number of years. This approach
requires us to determine values for model parameters (e.g.~population
variability). As an example, we can take the following population model
for population size \(N\) at time \(t\):

@@

where \(\epsilon\) is a normally-distributed random noise term with mean
\(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\) is
also the trend strength of the increase or decrease (i.e.~the rate of
increase). It is important to note that any population model could be
substituted for equation \ref{sim_model}, as in the supplementary
material (Figs. \ref{fig:min_time_dist_log_pop},
\ref{fig:shark_example}).

Statistical power is then the proportion of simulations that meet some
criteria. Here, our criteria is whether the slope parameter from linear
regression is significant at the 0.05 threshold with statistical power
of 0.8. Statistical power of 0.8 would indicate that, if there was a
true trend in abundance, there would be a 0.8 probability of detecting
the trend. Values of 0.05 for the significance level and 0.8 statistical
power are historical and it is important to examine the effect of
changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (\(t=40\)). It is
clear that statistical power increases quickly with increases in length
of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is
greater than 0.8 (the dotted line), that is the minimum time required
(\(T_{min}\)) to be confident in the detection of a long-term trend in
abundance.

@@

I use a database of 2444 population time series compiled in (Keith et
al. 2015); they compared the predictability of growth rates among
populations. The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010) and several other
sources (Keith et al. 2015). I filtered out short time series (less than
35 years), and those with missing data, leaving 822 time series. The
data includes information on 477 vertebrate species with a focus on
mammals, birds, and fish. The data also includes information on
generation length and survey specifications. For each time series, I
also calculate other variables of interest: coefficient of variance in
population size, long-term trend in abundance (slope coefficient from
simple linear regression), and temporal autocorrelation. All analyses
were conducted in R (R Core Team 2016).

For a subset of populations (\(n =\) 547), there is information on
life-history traits available from another paper (Myhrvold et al. 2015),
including body size and generation length. All 547 populations were
birds. I examine how the minimum time required is related to these
life-history traits (Fig. \ref{fig:biological_correlates}).

@@

I assume that each time series is long enough to include all necessary
information (e.g.~variability) about the population. In other words,
each time series is a representative sample. I first take all possible
contiguous subsamples of each time series. For example, a time series of
35 years would have 34 possible contiguous subsamples of length 2, 34
possible contiguous subsamples of length 3, and continuing until 1
possible contiguous subsample of length 35 (Gerber, DeMaster, and
Kareiva 1999; Giron-Nava et al. 2017). Next, I run a linear regression
for each subsampled time series. Then, I determine the proportion of
subsamples of a particular length that have estimated slope coefficients
which are statistically different from zero. I only look at the
proportion of samples where the long-term, or ``true'', time series also
has a significant slope. This proportion is a measure of statistical
power. Lastly, I determine which subsample length is required to achieve
a certain threshold of statistical power (0.8, Cohen (1992)). The
minimum subsampled length that met these criteria is the minimum time
series length required (\(T_{min}\)).

In the supplementary material, I show how the same approach described
here for more complicated population models. I also determine the
minimum time required to estimate long-term trends according to
generalized additive models, instead of the simple linear models used
here (Fig. \ref{fig:gam_example}).

@@

I determined the minimum time series length (\(T_{min}\)) required to
address a particular question of interest. What is the minimum time
series length required to determine, via linear regression, the
long-term population trend? Here, the minimum time series length
required had high enough statistical power (greater than 0.8) for a set
significance level (\(\alpha\)) of 0.05. It is also possible to alter
statistical power and \(\alpha\). Predictably, with increased
statistical power or decreased \(\alpha\), \(T_{min}\) increased (Fig.
\ref{fig:min_time_vs_alpha_beta}). I then estimated \(T_{min}\) using
two approaches. I briefly describe results from the simulation approach
and then discuss the empirical approach.

@@

I constructed a general population model where the trend strength
(i.e.~slope coefficient) over time could be a model parameter. I then
simulated time series of different lengths. From these simulations I
determined the minimum time series length required to achieve a certain
level of statistical power. In line with past work (Gerrodette 1987), I
found the \(T_{min}\) increases (i.e.~more time is required) with
decreases in trend strength and with increases in population variability
(Figs. \ref{fig:theoretical_approach}c,d).

I chose a simple model, but any other population model could be used
(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
specific model choice should be tailored to the population of interest.
I explored how the simulation approach could be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
Specifically, I determined the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). I found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
the simulation approach described above, the minimum time required for
the lemon shark population was strongly dependent on model parameters.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{(a) Distributions of the minimum time required for populations
from four different biological classes. (b) Distribution of minimum time
required for all populations regardless of biological class. The minimum
time required calculation corresponds to a significance level of 0.05
and statistical power of 0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) coefficient of variation in
interannual population size, and (c) temporal lag-1
autocorrelation.\label{fig:correlates}}
\end{figure}

I examined a database of 822 separate population time series
representing 477 species. This database consisted of vertebrate species
with a variety of life-history characteristics (Fig.
\ref{fig:biological_correlates}). I limited analyses to populations with
at least 35 years of continuous sampling. I then examined the minimum
time required to estimate long-term trends via linear regression.

Across all the populations, I found an average minimum time series
length required (\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution
(Fig. \ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
Actinopterygii) typically had estimates of \(T_{min}\) over 20 years.
Birds (class Aves) had a much wider distribution of \(T_{min}\), but
usually required less years of sampling. Differences between these
classes can be explained by differences in variability in population
size and strength of trends in abundance (Fig. \ref{fig:class}).

\subsubsection{Corrrelates for minimum time
required}\label{corrrelates-for-minimum-time-required}

The minimum time series length required was strongly correlated with
trend strength (i.e.~estimated slope coefficient from linear
regression), coefficient of variation in population size, and
autocorrelation in population size (Fig. \ref{fig:correlates}). This is
in line based on simulations here and those of others (Rhodes and Jonzen
2011). Using a generalized linear model, with a Poisson error structure,
all three of these explanatory variables were significant and had large
effect sizes (see Table A1). Combined, trend strength, coefficient of
variation in population, and autocorrelation account for 75.1\% of the
explained deviance (Zuur et al. 2009) in minimum time series length
required.

For a subset of the populations I combined time series data with a data
on life-history characteristics of amniotes (Myhrvold et al. 2015).
There was life-history information available for 547 populations
representing 315 different species, all of which were birds (Aves
class).

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) log adult body mass (grams), (d) maximum longevity
(years), and (e) incubation (days). The lines in each plot represent the
best fit line from linear regression.\label{fig:biological_correlates}}
\end{figure}

Some life-history traits were significant predictors for the minimum
time required (Fig. \ref{fig:biological_correlates}, Tables A2,A3).
However, none of these life-history traits explained a large part of the
variation in minimum time required. In a generalized linear model, all
of the life-history traits described in figure
\ref{fig:biological_correlates} account for only 5.99\% of the explained
deviance in minimum time series length required. In addition, when
accounting for trend strength, coefficient of variation, and
autocorrelation, no life-history traits were significant predictors of
the minimum time required (Table A3).

\subsubsection{Evalulating the IUCN
criteria}\label{evalulating-the-iucn-criteria}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Minimum time required to achieve 0.8 statistical power versus
the minimum time required under IUCN criteria A2 to classify a species
as vulnerable. Each point represents a single population, all of which
saw declines of 30\% or greater over a 10 year
period.\label{fig:IUCN_analysis}}
\end{figure}

The IUCN Red List Categories and Criteria suggest, under Criterion A2, a
species qualifies as vulnerable if it has experienced a 30\% decline
over 10 years, or 3 generations (whichever is longer) (IUCN 2012). I
examined a subset of populations with observed declines of 30\% or
greater over 10 years, qualifying all of them as vulnerable. This
resulted in n = 162 populations of fish, birds, and a single mammal. I
then compared the minimum time required to achieve 0.8 statistical power
(\(T_{min}\)) to the minimum time required under the IUCN criteria (Fig.
\ref{fig:IUCN_analysis}). For populations below the identity line in
figure \ref{fig:IUCN_analysis}, IUCN criteria would require more
sampling compared to estimates for \(T_{min}\). Further, populations
above the identity line are cases where the IUCN criteria would classify
a population as vulnerable despite not having sampled enough years to
achieve high statistical power (Fig. \ref{fig:IUCN_analysis}). The
silhouettes on figure \ref{fig:IUCN_analysis} highlight that species
with longer generation times typically have larger discrepancies between
\(T_{min}\) and the minimum time required for IUCN assessments (Fig.
\ref{fig:IUCN_correlates}).

\subsubsection{Sensitivity analysis}\label{sensitivity-analysis}

Lastly, I tested model sensitivity by using generalized additive models
(GAMs) instead of simple linear regression. Again, I examined the
minimum time required to estimate long-term population trends (Fig.
\ref{fig:gam_example}). I found that although I obtain a similar
distribution of minimum times required for GAMs, the minimum time
required for GAMs is on average shorter than for linear regression (Fig.
\ref{fig:min_time_dist_gam}).

@@

I explored two approaches to estimate the minimum time series length
required to address a particular question of interest. I asked, what is
the minimum time series length required to determine long-term
population trends using linear regression? This is one of the simplest
questions one could ask of a time series. The simulation-based approach
has been suggested by others, especially in situations more complicated
than that suited for classic power analysis (Gerrodette 1987; P. C.
Johnson et al. 2015; Bolker 2008). My simulations support past work that
longer time series are needed when the trend strength (i.e.~rate of
increase or decrease) is weak or when population variability is high
(Gerrodette 1987). I also showed how the simulation model can be altered
for a particular population (Fig. \ref{fig:shark_example}) or question
(Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).

Here, I focus on an empirical approach to estimate the minimum time
series length required to assess changes in abundance over time. I
examined 822 population time series (all longer than 35 years). I then
subsampled each to determine the minimum time required to achieve a
desired significance level and power for linear regression. Statistical
power is important as it provides on information as to the necessary
samples required to determine a significant trend (Legg and Nagy 2006).
I found that on average 15.91 years of continuous monitoring were
typically necessary (Fig. \ref{fig:min_time_dist}b). However, the
distribution of minimum time required was wide. This time-frame is in
line with past work on a short-lived snail species (Rueda-Cediel et al.
2015) and a long-lived whale species (Gerber, DeMaster, and Kareiva
1999). Hatch (2003) used seabird monitoring data to estimate minimum
sampling requirements. He found that the time required ranged from 11 to
69 years depending on species, trend strength, and study design.
